---
title: "Cost of Weather in Lives and Dollars"
author: "Jerad Acosta"
date: "October 26, 2014"
output: html_document
---
## The Cost of Weather in Lives and Dollars

### Synopsis
The data analyzed herein comes from the NCDC who collects if from the National
Weather Service which collects and receives the data from a variety of sources.

This data contains myriad information on weather data from accross the US.
This includes but is not limited to: locations, type of weather, time, as well
as tides and other natural phenomina.

Of greatest significance to this report is the collection of data on the impact
weather has on both the populations health and the economy as a whole.

To address this question variables such as Fatalities and Injuries were collected
and organized according to source and magnitude.

Additionally, damages to property, required storm preperation and other factors
such as crop destruction and related costs from delayed processes.


### Data Processing
Load Data and libraries
```{r }
library(plyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(grid)
library(gridExtra)
data <- read.csv("repdata-data-StormData.csv")

```

## Search for significant factors of damage to population health
**Subset data to contrain only factors of interest with respect to population health**
```{r }
health <- select(data, EVTYPE, FATALITIES, INJURIES)
```

Find range in Fatalities and Injuries to deduce most significant events
```{r}
range(health$FATALITIES)
sum(health$FATALITIES)
range(health$INJURIES)
sum(health$INJURIES)
```

In our 902297 observations we find 15145 fatalities ranging from 0 to 583
and 140528 injuries ranging 0 to 1700
Now lets filter to most signiticant Events with respect to Fatalities
To avoid confounding loss of data from events that may not cause massive
damage at once but over time let's include all events which results in at least
```{r}
healthmelt <- melt(health, id=c("EVTYPE"), measure.vars=c("FATALITIES","INJURIES"))
healthsum <- dcast(healthmelt, EVTYPE ~ variable, sum)
```

Now Arrange and filter out data with no
```{r}
top.fatal <- arrange(healthsum, desc(FATALITIES))
top.fatal <- filter(top.fatal, FATALITIES > 1)
top.injure <- arrange(healthsum, desc(INJURIES))
top.injure <- filter(top.injure, INJURIES > 1)
```

**Use graph to see how data is distributed accross categorical data such as Weather Events**
```{r}
par(mfcol = c(2,2), oma = c(0,0,2,0))
barplot(top.fatal$FATALITIES, xlab = "Weather Event", ylab = "Fatalities")
barplot(top.fatal$FATALITIES, ylab = "Injuries",
        sub = "Highest Ranked fatalities subset")
barplot(top.fatal$INJURIES, xlab = "Weather Event", ylab = "Injuries")
barplot(top.fatal$FATALITIES, ylab = "Fatalities",
        sub = "Highest Ranked Injuries subset")
mtext("Grid of Worst Weather Events by ranked and plotted by Injury and Fatality count")
par(mfrow = c(1,1))
```

From the data here we can see that the majority of the damage is done by
a few significant Weather Event Types.
Too look Closer at them we can subset and graph with labels

```{r two column ggplot2, fig.width= 16}
# Top Fatal Weather Events
fat <- qplot(EVTYPE, FATALITIES, data = top.fatal[1:7,], fill = EVTYPE,
             geom = "bar", stat="identity", main = "Top Causes of Death")
# Top Injuries Weather Events
inj <- qplot(EVTYPE, INJURIES, data = top.injure[1:7,], fill = EVTYPE,
             geom = "bar", stat="identity", main = "Top Cuase of Injury")
grid.arrange(fat, inj, ncol = 2, main = "Damage to Health By Weather Event")
```


### Results
From this we can conclude that the most devistating of all weather events
to the health of citizens in the United States is by and Far Tornados.
So much so, that it is on a different scale of magnitude.

It is also worth noting that Excessive Heat shows up quite significantly
and that floods show up in two forms of *flooding* and *flash flooding*